### 2024-07-07 causal-inference.qmd
20240706 - clarifications
20240706 - bidirectional effects

- `X` causes `Y`
- `Y` causes `X`
- a third variable (i.e., [confound](#sec-causalDiagramConfounding)), `Z`, influences both `X` and `Y`
- the association between `X` and `Y` is spurious

@@

- Two variables (`A` and `B`) influence a [collider](#sec-colliderBias) (`M`).
- The [collider](#sec-colliderBias) influences two variables, `X` and `Y`.
- `A` also influences `X`.
- `B` also influences `Y`.
- `A` and `B` are not causally related
- `X` and `Y` are not causally related

>>>>

- `X` causes `Y`
- `Y` causes `X`
- `X` and `Y` are bidirectional: `X` causes `Y` and `Y` causes `X`
- a third variable (i.e., [confound](#sec-causalDiagramConfounding)), `Z`, influences both `X` and `Y`
- the association between `X` and `Y` is spurious

@@

- Two variables (`A` and `B`) influence a [collider](#sec-colliderBias) (`M`).
- The [collider](#sec-colliderBias) influences two variables, `X` and `Y`.
- `A` also influences `X`.
- `B` also influences `Y`.
- `A` and `B` are not causally related.
- `X` and `Y` are not causally related.

### 2024-07-07 research-methods.qmd
20240706 - bidirectional effects

- `X` causes `Y`
- `Y` causes `X`
- a third variable (i.e., confound), `Z`, influences both `X` and `Y`
- the association between `X` and `Y` is spurious

@@

A third possibility is that the association could reflect the influence of a third variable.
If a third variable is a common cause of each and accounts for their association, it is a *confound*.
An observed association between `X` and `Y` could reflect a confound—i.e., a cause (`Z`) that influences both `X` and `Y`, which explains why `X` and `Y` are correlated even though they are not causally related.
A third variable confound that is a common cause of both `X` and `Y` is depicted in @fig-correlationAndCausation3.
For instance, it may not be that sport drink consumption per se influences player performance; rather, it may be that players who are more intelligent or have more financial resources tend to drink more sports drinks and also tend to perform better.
In this case, intelligence or financial resources may be a confound that influences both sports drink consumption and player performance, but sports drink consumptions—though correlated with player performance—does not influence player performance.

>>>>

- `X` causes `Y`
- `Y` causes `X`
- `X` and `Y` are bidirectional: `X` causes `Y` and `Y` causes `X`
- a third variable (i.e., confound), `Z`, influences both `X` and `Y`
- the association between `X` and `Y` is spurious

@@

A third possibility is that the association reflects a bidirectional effect, where `X` causes `Y` and `Y` causes `X`, as depicted in INSERT.
For instance, consumption of more sports drink may improve players' performance, and greater performance in turn may lead players to drink more sports drink.

A fourth possibility is that the association could reflect the influence of a third variable.
If a third variable is a common cause of each and accounts for their association, it is a *confound*.
An observed association between `X` and `Y` could reflect a confound—i.e., a cause (`Z`) that influences both `X` and `Y`, which explains why `X` and `Y` are correlated even though they are not causally related.
A third variable confound that is a common cause of both `X` and `Y` is depicted in @fig-correlationAndCausation3.
For instance, it may not be that sport drink consumption per se influences player performance; rather, it may be that players who are more intelligent or have more financial resources tend to drink more sports drinks and also tend to perform better.
In this case, intelligence or financial resources may be a confound that influences both sports drink consumption and player performance, but sports drink consumptions—though correlated with player performance—does not influence player performance.
